\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3.2\baselineskip}

### 2017-03-27 paper/ms.tex

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are popular stochastic
models for time-series analysis.
For GP modeling, a functional form is chosen to describe the autocovariance
of the data and the parameters of this function are fit for or marginalized.
In the astrophysical literature, GPs have been used to model stochastic
variability in light curves of stars \citep{Brewer:2009}, active galactic
nuclei \citep{Kelly:2014}, and the logarithmic flux of X-ray binaries
\citep{Uttley:2005}.
They have also been used as models for the cosmic microwave background
\citep{Bond:1987, Bond:1999, Wandelt:2003}, correlated instrumental noise
\citep{Gibson:2012}, and spectroscopic calibration \citep{Czekala:2017,
Evans:2015}.
While these models are widely applicable, their use has been limited, in
practice, by the computational cost and scaling.
In general, the cost of computing a GP likelihood scales as the cube of
the number of data points $\mathcal{O}(N^3)$ and in the current era of large
time domain surveys~--~with as many as $\sim10^{4-9}$ targets with
$\sim10^{3-5}$ observations each~---~this scaling is prohibitive.

@@

GP are stochastic models consisting of a mean function
$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ parameterized by
$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, the log-likelihood of observing a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.

@@

By rewriting the exponentials in \eq{celerite-kernel-complex} as sums of sine
and cosine functions, we can see the autocorrelation structure is defined by a
mixture of quasiperiodic oscillators
\begin{eqnarray}\eqlabel{celerite-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
For clarity, we refer to the argument within the sum as a ``\celeriteterm\
term'' for the remainder of this paper.
The kernel function in \eq{celerite-kernel} is qualitatively similar to the
The Fourier transform\footnote{Here and throughout we have defined the Fourier
transform of the function $f(t)$ as $F(\omega)={(2\,\pi)}^{-1/2}\,
\int_{-\infty}^\infty f(t)\,e^{i\,\omega\,t}\dd t$.} of this covariance
function is the power spectral density (PSD) of the process and it is given by
\begin{eqnarray}\eqlabel{celerite-psd}
S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
return to a more general discussion shortly but we start by considering some
special cases.

@@

To provide some insight for this method, we follow
\citet{Ambikasaran:2015} and start by working through a simple example.
In this case, we assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This is the result of setting $J=1$, $b=0$, and $d=0$ in
\eq{celerite-kernel-complex} and is the model studied by
\citet{Rybicki:1995}.
To demonstrate that the likelihood of this model can be computed in
$\mathcal{O}(N)$, we write out the full system of
equations that must be solved to apply the inverse of $K$ and compute
the first term of \eq{gp-likelihood}.
In matrix notation, this is
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{z} &=& \bvec{y}, \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ z_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix}
\end{eqnarray}
where our goal is to solve for the unknown vector \bvec{z} for a given matrix
$K$ and vector \bvec{y}.
In \eq{impl-matrix}, we have assumed that the mean function is zero but a
non-zero mean could be included by replacing \bvec{y} by
$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
Now, if we introduce the variables
\begin{eqnarray}\eqlabel{algo-first}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
Rewriting the system defined by \eq{algo-first} through \eq{algo-last}
as a matrix equation shows the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
Following \citet{Ambikasaran:2015} we call this the ``extended'' system and
rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ we started
with, it is now sparse with banded structure that can be exploited to solve
the system efficiently.
In particular, sparse solvers are available that can perform a
LU-decomposition of matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can use these algorithms to solve our system
exactly because the target vector $\bvec{z}$ is a subset of the elements of
$\bvec{z}_\mathrm{ext}$.

@@

The extended system defined in the previous section is sparse with typically fewer than
a few percent non-zero entries and band structure.
In this section, we empirically investigate the performance and scaling
of three different algorithms for solving this extended system:

@@

The benchmark experiments shown here were performed on a MacBook Pro with two
2.6~GHz CPUs but we find similar results on a Dell workstation with 16 2.7~GHz
CPUs and running Ubuntu.
\Figure{benchmark} shows how the cost of computing \eq{gp-likelihood} scales
with $N$ and $J$ using the \texttt{vanilla} solver.
As expected theoretically, the scaling is linear in $N$ for all $N$ and cubic
in $J$ for large $J$.
\Figure{benchmark-lapack} and \Figure{benchmark-sparse} are the same plots for
the \texttt{lapack} and \texttt{sparse} solvers respectively.
For each of these optimized solvers, the empirical scaling is
$\mathcal{O}(N\,J^2)$ at the cost of some extra overhead.
Therefore, for $J \lesssim 5$ or $10$ the \texttt{vanilla} solver is more
efficient than the other algorithms.
The real world performance of \celerite\ depends on the specific platform,
hardware, and \project{LAPACK}/\project{BLAS} implementation but we have found
qualitatively similar results across popular platforms and state-of-the-art
libraries.

@@

These special cases demonstrate that the stochastically-driven simple harmonic
oscillator provides a physically motivated model that is flexible enough to
describe a wide range of stellar variations.
Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
model for asteroseismic oscillations.
In practice, we take a sum over oscillators with different values of $Q$,
$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
stellar time series.
Since this kernel is exactly described by the exponential kernel, the
likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
measurements in $\mathcal{O}(N)$ operations using the \celerite\ method
described in the previous section.

@@

For this example, we simulate a dataset using a known GP model with a kernel
outside of the support of a \celeriteterm\ process.
This means the true autocorrelation of the process can never be correctly
represented by the model that we're using to fit but we use this example
to demonstrate that, at least in this case, valid inferences can still be made
about the physical parameters of the model.

For this example the data are simulated from a quasiperiodic GP with the
kernel
\begin{eqnarray}\eqlabel{sim-wrong-true}
k_\mathrm{true} (\tau) = \alpha\,
    \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{true}}\right)
\end{eqnarray}
where $P_\mathrm{true}$ is the fundamental period of the process.
This autocorrelation structure corresponds to the power spectrum
\begin{eqnarray}
S_\mathrm{true} (\omega) = \frac{\lambda\,\alpha}{2}\,\left[
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right) +
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega+
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right)
\right]
\end{eqnarray}
which, for large values of $\omega$, falls off exponentially.
When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
$\omega^{-4}$ at most~--~it is clear that a \celeriteterm\ model can never
perfectly reproduce the structure of this process.
That being said, we demonstrate that rigorous inferences can be made
about $P_\mathrm{true}$ even with an effective model.
The left panel of \Figure{simulated-wrong} shows the simulated dataset.
We then fit this simulated data using the product of two SHO terms
(\eqalt{sho-kernel}) where one of the terms has $S_0 = 1$ and $Q =
1/\sqrt{2}$ and the other has $\omega_0 = 2\,\pi/P$.
We note that using \eq{product-rule}, the product of two \celeriteterm\ terms can
also be expressed using \celerite.
We apply log-uniform priors on all the parameters and use \emcee\ to sample
the joint posterior probability for all of the parameters.
The inferred distribution for the parameter $P$ is shown in the right panel of
\Figure{simulated-wrong} and compared to the true period $P_\mathrm{true}$ and
the inferences made using the correct model (\eqalt{sim-wrong-true}).
The inference made using this effective \celeriteterm\ model are indistinguishable
from the inferences made using the correct model but substantially less
computation time is required for the \celerite\ inference.

@@

\celerite\ alleviates these problems by providing a physically motivated
probabilistic model that can be evaluated efficiently even for large datasets.
In practice, one would model the star as a mixture of stochastically-driven
simple harmonic oscillators where the amplitudes and frequencies of the
oscillations are computed using a physical model and evaluate the probability
of the observed time series using a GP where the PSD is a sum of terms given
by \eq{sho-psd}.
This gives us a method of computing the likelihood function for the parameters
of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta \nu$, or
other, more fundamental parameters) \emph{conditioned on the observed time
series} in $\mathcal{O}(N)$ operations.
In other words, \celerite\ provides a computationally efficient framework that
can be combined with physically-motivated models of stars and numerical
inference methods to make rigorous probabilistic measurements of asteroseismic
parameters in the time domain.
We expect that this has the potential to push asteroseismic analysis to lower
signal-to-noise datasets and we hope to revisit this idea in a subsequent
paper.

To demonstrate the method, we use a very simple heuristic model where the PSD
is given by a mixture of 8 components with amplitudes and frequencies
specified by $\nu_\mathrm{max}$, $\Delta \nu$, and several nuisance
parameters.
The first term is used to capture the granulation ``background''
\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
$S_g$ and $\omega_g$.
The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
parameter shared between terms and the frequencies are given by
\begin{eqnarray}
\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
\end{eqnarray}
and the amplitudes are given by
\begin{eqnarray}
S_{0,\,j} =
    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
\end{eqnarray}
where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
are shared nuisance parameters.
This model could be easily extended to include small frequency splitting and
$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by the fundamental
physical parameters of the star.

@@

As discussed previously, the \celerite\ method is based upon a method
developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
In the case where the kernel function has a single term of the form
\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
and the Rybicki \& Press method.
Their method does not require solving the extended matrix because the matrix
inverse can be calculated analytically but the method is made slightly more
complicated when white noise is included.
We implement their full method and find empirically that it is more
computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$
but we note that the requirement of a single, real, exponential kernel is
extremely restrictive and this model cannot describe quasi-periodic
variability.

>>>>

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are popular stochastic
models for time-series analysis.
For GP modeling, a functional form is chosen to describe the autocovariance
of the data and the parameters of this function are fit for or marginalized.
In the astrophysical literature, GPs have been used to model stochastic
variability in light curves of stars \citep{Brewer:2009}, active galactic
nuclei \citep{Kelly:2014}, and the logarithmic flux of X-ray binaries
\citep{Uttley:2005}.
They have also been used as models for the cosmic microwave background
\citep{Bond:1987, Bond:1999, Wandelt:2003}, correlated instrumental noise
\citep{Gibson:2012}, and spectroscopic calibration \citep{Czekala:2017,
Evans:2015}.
While these models are widely applicable, their use has been limited, in
practice, by the computational cost and scaling.
In general, the cost of computing a GP likelihood scales as the cube of
the number of data points $\mathcal{O}(N^3)$ and in the current era of large
time domain surveys~--~with as many as $\sim10^{4-9}$ targets with
$\sim10^{3-5}$ observations each~--~this scaling is prohibitive.

@@

GPs are stochastic models consisting of a mean function
$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ parameterized by
$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, the log-likelihood of observing a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.

@@

By rewriting the exponentials in \eq{celerite-kernel-complex} as sums of sine
and cosine functions, we can see the autocorrelation structure is defined by a
mixture of quasiperiodic oscillators
\begin{eqnarray}\eqlabel{celerite-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
For clarity, we refer to the argument within the sum as a ``\celeriteterm\
term'' for the remainder of this paper.
The Fourier transform\footnote{Here and throughout we have defined the Fourier
transform of the function $f(t)$ as $F(\omega)={(2\,\pi)}^{-1/2}\,
\int_{-\infty}^\infty f(t)\,e^{i\,\omega\,t}\dd t$.} of this covariance
function is the power spectral density (PSD) of the process and it is given by
\begin{eqnarray}\eqlabel{celerite-psd}
S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
return to a more general discussion shortly but we start by considering some
special cases.

@@

To provide some insight for this method, we follow
\citet{Ambikasaran:2015} and start by working through a simple example.
In this case, we assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This is the result of setting $J=1$, $b=0$, and $d=0$ in
\eq{celerite-kernel-complex} and it is the model studied by
\citet{Rybicki:1995}.
To demonstrate that the likelihood of this model can be computed in
$\mathcal{O}(N)$, we write out the full system of
equations that must be solved to apply the inverse of $K$ and compute
the first term of \eq{gp-likelihood}.
In matrix notation, this is
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{z} &=& \bvec{y}, \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ z_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix}
\end{eqnarray}
where our goal is to solve for the unknown vector \bvec{z} for a given matrix
$K$ and vector \bvec{y}.
In \eq{impl-matrix}, we have assumed that the mean function is zero but a
non-zero mean could be included by replacing \bvec{y} by
$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
Now, if we introduce the variables
\begin{eqnarray}\eqlabel{algo-first}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
Rewriting the system defined by \eq{algo-first} through \eq{algo-last}
as a matrix equation shows the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
Following \citet{Ambikasaran:2015} we call this the ``extended'' system and
rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ we started
with, it is now sparse with banded structure that can be exploited to solve
the system efficiently.
In particular, sparse solvers are available that can perform a
LU-decomposition of matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can use these algorithms to solve our system
exactly because the target vector $\bvec{z}$ is a subset of the elements of
$\bvec{z}_\mathrm{ext}$.

@@

The extended system defined in the previous section is sparse, with typically
fewer than a few percent non-zero entries and band structure.
In this section, we empirically investigate the performance and scaling
of three different algorithms for solving this extended system:

@@

The benchmark experiments shown here were performed on a MacBook Pro with two
2.6~GHz CPUs but we find similar results on a Dell workstation with 16 2.7~GHz
CPUs and running Ubuntu.
\Figure{benchmark} shows how the cost of computing \eq{gp-likelihood} scales
with $N$ and $J$ using the \texttt{vanilla} solver.
As expected theoretically, the scaling is linear in $N$ for all $N$ and cubic
in $J$ for large $J$.
\Figure{benchmark-lapack} and \Figure{benchmark-sparse} are the same plots for
the \texttt{lapack} and \texttt{sparse} solvers respectively.
For each of these optimized solvers, the empirical scaling is
$\mathcal{O}(N\,J^2)$ at the cost of some extra overhead.
Therefore, for $J \lesssim 5$ or $10$ the \texttt{vanilla} solver is more
efficient than the other algorithms.
The real world performance of \celerite\ depends on the specific platform,
hardware, and \project{LAPACK}/\project{BLAS} implementation, but we have
found qualitatively similar results across popular platforms and
state-of-the-art libraries.

@@

These special cases demonstrate that the stochastically-driven simple harmonic
oscillator provides a physically motivated model that is flexible enough to
describe a wide range of stellar variations.
Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
model for asteroseismic oscillations.
In practice, we take a sum over oscillators with different values of $Q$,
$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
stellar time series.
Since this kernel is exactly described by the exponential kernel, the
likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
measurements in $\mathcal{O}(N)$ operations using the \celeriteterm\ method
described in the previous section.

@@

For this example, we simulate a dataset using a known GP model with a kernel
outside of the support of a \celeriteterm\ process.
This means the true autocorrelation of the process can never be correctly
represented by the model that we are using to fit, but we use this example
to demonstrate that, at least in this case, valid inferences can still be made
about the physical parameters of the model.

For this example, the data are simulated from a quasiperiodic GP with the
kernel
\begin{eqnarray}\eqlabel{sim-wrong-true}
k_\mathrm{true} (\tau) = \alpha\,
    \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{true}}\right)
\end{eqnarray}
where $P_\mathrm{true}$ is the fundamental period of the process.
This autocorrelation structure corresponds to the power spectrum
\begin{eqnarray}
S_\mathrm{true} (\omega) = \frac{\lambda\,\alpha}{2}\,\left[
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right) +
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega+
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right)
\right]
\end{eqnarray}
which, for large values of $\omega$, falls off exponentially.
When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
$\omega^{-4}$ at most~--~it is clear that a \celeriteterm\ model can never
perfectly reproduce the structure of this process.
That being said, we demonstrate that rigorous inferences can be made
about $P_\mathrm{true}$ even with an effective model.
The left panel of \Figure{simulated-wrong} shows the simulated dataset.
We then fit this simulated data using the product of two SHO terms
(\eqalt{sho-kernel}) where one of the terms has $S_0 = 1$ and $Q =
1/\sqrt{2}$ and the other has $\omega_0 = 2\,\pi/P$.
We note that using \eq{product-rule}, the product of two \celeriteterm\ terms can
also be expressed using \celerite.
We apply log-uniform priors on all the parameters and use \emcee\ to sample
the joint posterior probability for all of the parameters.
The inferred distribution for the parameter $P$ is shown in the right panel of
\Figure{simulated-wrong} and compared to the true period $P_\mathrm{true}$ and
the inferences made using the correct model (\eqalt{sim-wrong-true}).
The inference made using this effective \celeriteterm\ model are indistinguishable
from the inferences made using the correct model and substantially less
computation time is required for the \celerite\ inference.

@@

\celerite\ alleviates these problems by providing a physically motivated
probabilistic model that can be evaluated efficiently even for large datasets.
In practice, one would model the star as a mixture of stochastically-driven
simple harmonic oscillators where the amplitudes and frequencies of the
oscillations are computed using a physical model and evaluate the probability
of the observed time series using a GP where the PSD is a sum of terms given
by \eq{sho-psd}.
This gives us a method of computing the likelihood function for the parameters
of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta \nu$, or
other more fundamental parameters) \emph{conditioned on the observed time
series} in $\mathcal{O}(N)$ operations.
In other words, \celerite\ provides a computationally efficient framework that
can be combined with physically-motivated models of stars and numerical
inference methods to make rigorous probabilistic measurements of asteroseismic
parameters in the time domain.
We expect that this has the potential to push asteroseismic analysis to lower
signal-to-noise datasets and we hope to revisit this idea in a subsequent
paper.

To demonstrate the method, we use a very simple heuristic model where the PSD
is given by a mixture of 8 components with amplitudes and frequencies
specified by $\nu_\mathrm{max}$, $\Delta \nu$, and several nuisance
parameters.
The first term is used to capture the granulation ``background''
\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
$S_g$ and $\omega_g$.
The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
parameter shared between terms and the frequencies are given by
\begin{eqnarray}
\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
\end{eqnarray}
and the amplitudes are given by
\begin{eqnarray}
S_{0,\,j} =
    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
\end{eqnarray}
where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
are shared nuisance parameters.
This model could be extended to include more angular modes and
$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by the fundamental
physical parameters of the star.

@@

As discussed previously, the \celerite\ method is based upon a method
developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
In the case where the kernel function has a single term of the form given in
\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
and the Rybicki \& Press method.
Their method does not require solving the extended matrix because the matrix
inverse can be calculated analytically but the method is made slightly more
complicated when white noise is included.
We implement their full method and find empirically that it is more
computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$
but we note that the requirement of a single, real, exponential kernel is
extremely restrictive and this model cannot describe quasi-periodic
variability.

Another popular method uses the fact that, in the limit of evenly-spaced data
and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
\citep[for example][]{Dillon:2013}.
There are exact methods for solving Toeplitz matrix equations that scale as
$\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N)$
\citep{Wilson:2014}.
The Toeplitz method is, in some ways, more flexible than \celerite\ because it
can be used with any stationary kernel, but it requires uniformly spaced data
and the scaling is worse than \celerite\ so it is less efficient when applied
to large datasets.
